Uncertainty propagation for predictor effects and bias correction in generalized linear (mixed) models

### 2022-01-05 variable_predictions.Rnw
Autocommit

\newcommand{\bX}{{\mathbf X}}
\newcommand{\bbeta}{{\boldsymbol \beta}}
\newcommand{\boldeta}{{\boldsymbol \eta}}
\newcommand{\boldmu}{{\boldsymbol \mu}}

@@

set.seed(9991)
load("variable_predictions_objs.rda")
load("justify_preds.rda")
load("cubic_predictors_preds.rda")
load("cubic_predictors_preds_adjust.rda")
load("glm_one_predictor_preds.rda")
load("glme_random_intercept_preds.rda")
load("binom_multiple_outcomes_preds_plots.rda")
@

@@

In population-based approach, $\eta^\star$ is constructed as a function of properly constituted levels of focal predictors, $x_f$, and entire population of the non-focal predictors, $\nset{x}$, as opposed to averaged non-focal predictors, $\nset{{\bar{x}}}$, described in the previous section. More specifically:
%
\begin{itemize}
\item compute linear predictor associated with the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
\item compute linear predictor associated with the focal predictors, $\eta_{jf} = \sum{\beta_f x_{jf}}$
\item for every value of the focal predictor, $\eta_{jf}$:
\begin{itemize}
\item $\eta_j^\star  = \eta_{jf} + \nset{\eta}$
\item $\hat{y}_j  = \textrm{mean} ~ g^{-1} \left(\eta_j^\star\right)$
\end{itemize}
\end{itemize}
%
We make similar adjustments to compute the variances of the predictions at every level of the focal predictor:
%
\begin{align}
\sigma_{jf}^2 = \textrm{Diag}(\bX^\star_{jc} \Sigma \bX^{\star\top}_{jc})
\end{align}
%
where $\bX^{\star}_{jc} = \{\bX_{jf}^\star, \nset{{\bX}^\star} - \nset{{\bar{\bX}}^\star}\}$ and 
%
\begin{align}
\mathrm{CI}_j = \mathrm{mean} ~ \eta_j^\star \pm q\sigma_{jf}
\end{align}
%

@@

Consider model described in \ref{sim:lm_no_interaction} with only \code{Age} as the predictor such that 
%
\begin{align}\label{sim:glm_one_pred}
\mathrm{logit(status} = 1) &= \eta \nonumber\\
\mathrm{\eta} &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i + \epsilon_i
\end{align}
%
where \code{status} is a binary outcome with 0 for unimproved and 1 for improved water quality in the slums. By having only a single predictor, we may be able to investigate the contribution of nonlinear link function on the predicted probability of improved water quality.

@@

Suppose, in the simulation described in Section~\ref{sec:stats_background}, the observations are  recorded at least once per household and we are interested in the status improved water services (improved or unimproved). In particular, let $\mathrm{H}$ be the number of households indexed by the grouping factor, and $\mathrm{h}[i]$ be the household of the $i$th observation such that 
%
\begin{align}\label{sim:lme_one_pred}
\mathrm{logit(status} = 1) &= \eta \nonumber\\
\mathrm{\eta} &= \beta_0 + \alpha_{0,\mathrm{h}[i]} + \beta_{\mathrm{A}}\mathrm{Age}_i + \epsilon_i \nonumber\\
\alpha_{0,\mathrm{h}} &\sim \mathrm{Normal}(0, 5),~~ h=1, \cdots, \mathrm{10} \nonumber\\
\mathrm{Age}_i &\sim \mathrm{Normal}(0, 3) \nonumber\\
\epsilon_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
\beta_0 &= 5 \nonumber\\
\beta_{\mathrm{A}} &= 0.8 \nonumber\\
i &= 1,\cdots, 500
\end{align}
%
We compare predictor effects generated without any bias correction to those corrected for the bias, and then compare the average predicted probability to the observed. The results are shown in Figure~\ref{fig:glme_plots}. The uncorrected approach overestimates the predicted probabilities as opposed to the bias corrected (population-based approach) estimates which closely match the observations and the observed average probability. In the latter approach, the random effects are incorporated (as additional non-focal predictor) in computation of $\nset{\eta}$ to take into account the bias induced by the random effect predictor.
%
\begin{figure}[H]
\begin{center}
<<glme_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
print(glme_plots)
@
\end{center}
\caption{A comparison of bias corrected and uncorrected predictor effects. The red lines are the bias uncorrected estimates while the black ones are the bias corrected. The horizontal red and black lines are average predicted probability of improved water. The horizontal grey (overlaid by the black) dotted line is the observed average probability of improved water. The vertical dotted grey line is the average age while the grey points are the binned observations.}
\label{fig:glme_plots}
\end{figure}
%

@@

We now consider a bit more complicated model involving two outcome variables (services) -- status (improved or unimproved) of water and garbage services. In particular, we simulate an intercept only multivariate random effect model to, \emph{jointly}, predict probability of improved water and garbage services based on age and wealth index. Each household has its own correlated effect on the intercepts (random-intercept). In addition, we add latent variable to account for unobserved correlations.
%
\begin{align}\label{sim:lme_multi_outcome}
\mathrm{services}_{{\mathrm{h}[i]\mathrm{s}}} &= \mathrm{Binomial}(P_{{\mathrm{h}[i]\mathrm{s}}}) \nonumber\\
\mathrm{logit}(P_{{\mathrm{h}[i]\mathrm{s}}}) &= \beta_{0,s} + \alpha_{0,\mathrm{h}[i]\mathrm{s}} + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{Ls}}\mathrm{Latent}_{i\mathrm{s}} + \epsilon_{i\mathrm{s}} \nonumber\\
\alpha_{0,\mathrm{h}\mathrm{s}} &\sim \mathrm{Normal}(0, 2),~~ h=1, \cdots, \mathrm{H} \nonumber\\
\mathrm{Age}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
\mathrm{Latent}_{[i]\mathrm{s}} &\sim \mathrm{Normal}(0, 1) \nonumber\\
\epsilon_{[i]\mathrm{\text{\tiny{water}}}} &\sim \mathrm{Normal}(0, 2) \nonumber\\
\epsilon_{[i]\mathrm{\text{\tiny{garbage}}}} &\sim \mathrm{Normal}(0, 3) \nonumber\\
\beta_0 &= 5 \nonumber\\
\beta_{\mathrm{A}} &= 2 \nonumber\\
\beta_{\mathrm{W}} &= -3 \nonumber\\
\beta_{\mathrm{L\text{\tiny{water}}}} &= 1 \nonumber\\
\beta_{\mathrm{L\text{\tiny{garbage}}}} &= -5 \nonumber\\
i &= 1,\cdots, 100
\end{align}
%
Where:
\begin{itemize}
\item $\beta_{0,s}$ \textbf{is the service specific effect???}, i.e., baseline log odds of a household having improved service
\item $\alpha_{0,\mathrm{h}\mathrm{s}}$ household specific random intercepts, i.e., captures the offset from the overall baseline (mean)
\end{itemize}

@@

>>>>

\newcommand{\bX}{{\mathbf X}}
\newcommand{\bZ}{{\mathbf Z}}
\newcommand{\bbeta}{{\boldsymbol \beta}}
\newcommand{\boldeta}{{\boldsymbol \eta}}
\newcommand{\boldmu}{{\boldsymbol \mu}}

@@

set.seed(9991)
load("variable_predictions_objs.rda")
load("mediate_bin_preds.rda")
load("mediate_preds.rda")
load("justify_preds.rda")
load("cubic_predictors_preds.rda")
load("cubic_predictors_preds_adjust.rda")
load("glm_one_predictor_preds.rda")
load("glme_random_intercept_preds.rda")
load("binom_multiple_outcomes_preds_plots.rda")
@

@@

In population-based approach, $\eta^\star$ is constructed as a function of properly constituted levels of focal predictors, $x_f$, and entire population of the non-focal predictors, $\nset{x}$, as opposed to averaged non-focal predictors, $\nset{{\bar{x}}}$, described in the previous section. More specifically:
%
\begin{itemize}
\item compute linear predictor associated with the non-focal predictors, $\nset{\eta} = \sum \nset{\beta} \nset{x}$
\item compute linear predictor associated with the focal predictors, $\eta_{jf} = \sum{\beta_f x_{jf}}$
\item for every value of the focal predictor, $\eta_{jf}$:
%
\begin{align}\label{eq:pop_eta} 
\eta_j^\star  &= \eta_{jf} + \nset{\eta} \\
\hat{y}_j  &= \textrm{mean} ~ g^{-1} \left(\eta_j^\star\right)
\end{align}
%
\end{itemize}
%
We make similar adjustments to compute the variances of the predictions at every level of the focal predictor:
%
\begin{align}
\sigma_{jf}^2 = \textrm{Diag}(\bX^\star_{jc} \Sigma \bX^{\star\top}_{jc})
\end{align}
%
where $\bX^{\star}_{jc} = \{\bX_{jf}^\star, \nset{{\bX}^\star} - \nset{{\bar{\bX}}^\star}\}$ and 
%
\begin{align}
\mathrm{CI}_j = \mathrm{mean} ~ \eta_j^\star \pm q\sigma_{jf}
\end{align}
%

For models with random effects components, another of adjustments can be made to correct for bias induced by the random effects terms. In the population approach, we treat the random effects terms as additional non-focal predictors and simply make adjustment to Equation~\ref{eq:pop_eta}. In particular
%
\begin{align}\label{eq:pop_eta_re} 
\tau &= \bZ b \nonumber \\
\eta_j^\star  &= \eta_{jf} + \nset{\eta} + \tau_j
\end{align}
%
where $\bZ$ and $b$ are the design matrix and a vector of random effects, respectively.

@@

Consider model described in \ref{sim:lm_no_interaction} with only \code{Age} as the predictor such that 
%
\begin{align}\label{sim:glm_one_pred}
\mathrm{status}_i &\sim \mathrm{Binomial}(1, \mathrm{P(status}_i = 1)) \nonumber\\
\mathrm{logit(P(status}_i = 1)) &= \eta_i \nonumber\\
\mathrm{\eta}_i &= \beta_0 + \beta_{\mathrm{A}}\mathrm{Age}_i \nonumber\\
\mathrm{Age}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
\beta_0 &= 5 \nonumber\\
\beta_{\mathrm{A}} &= 0.8 \nonumber\\
i &= 1,\cdots, 500
\end{align}
%
where \code{status} is a binary outcome with 0 for unimproved and 1 for improved water quality in the slums. By having only a single predictor, we may be able to investigate the contribution of nonlinear link function on the predicted probability of improved water quality.

@@

Suppose, in the simulation described in Section~\ref{sec:stats_background}, the observations are  recorded at least once per household and we are interested in the status improved water services (improved or unimproved). In particular, let $\mathrm{H}$ be the number of households indexed by the grouping factor, and $\mathrm{h}[i]$ be the household of the $i$th observation such that 
%
\begin{align}\label{sim:lme_one_pred}
\mathrm{status}_i &\sim \mathrm{Binomial}(1, \mathrm{P(status}_i = 1)) \nonumber\\
\mathrm{logit(P(status}_i = 1)) &= \eta_i \nonumber\\
\mathrm{\eta}_i &= \beta_0 + \alpha_{0,\mathrm{h}[i]} + \beta_{\mathrm{A}}\mathrm{Age}_i \nonumber\\
\alpha_{0,\mathrm{h}} &\sim \mathrm{Normal}(0, 5),~~ h=1, \cdots, \mathrm{10} \nonumber\\
\mathrm{Age}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
\beta_0 &= 5 \nonumber\\
\beta_{\mathrm{A}} &= 0.8 \nonumber\\
i &= 1,\cdots, 500
\end{align}
%
We compare predictor effects generated without any bias correction to those corrected for the bias, and then compare the average predicted probability to the observed. The results are shown in Figure~\ref{fig:glme_plots}. The uncorrected approach overestimates the predicted probabilities as opposed to the bias corrected (population-based approach) estimates which closely match the observations and the observed average probability. In the latter approach, the random effects are incorporated (as additional non-focal predictor) in computation of $\nset{\eta}$ to take into account the bias induced by the random effect predictor.
%
\begin{figure}[H]
\begin{center}
<<glme_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
print(glme_plots)
@
\end{center}
\caption{A comparison of bias corrected and uncorrected predictor effects. The red lines are the bias uncorrected estimates while the black ones are the bias corrected. The horizontal red and black lines are average predicted probability of improved water. The horizontal grey (overlaid by the black) dotted line is the observed average probability of improved water. The vertical dotted grey line is the average age while the grey points are the binned observations.}
\label{fig:glme_plots}
\end{figure}
%

@@

We now consider a bit more complicated model involving two outcome variables (services) -- status (improved or unimproved) of water and garbage services. In particular, we simulate an intercept only multivariate random effect model to, \emph{jointly}, predict probability of improved water and garbage services based on age and wealth index. Each household has its own correlated effect on the intercepts (random-intercept). In addition, we add latent variable to account for unobserved correlations.
%
\begin{align}\label{sim:lme_multi_outcome}
\mathrm{status}_{{\mathrm{h}[i]\mathrm{s}}} &\sim \mathrm{Binomial}(1, P(\mathrm{status}_{{\mathrm{h}[i]\mathrm{s}}} = 1)) \nonumber\\
\mathrm{logit}(P(\mathrm{status}_{{\mathrm{h}[i]\mathrm{s}}} = 1)) &= \beta_{0,s} + \alpha_{0,\mathrm{h}[i]\mathrm{s}} + \beta_{\mathrm{A}}\mathrm{Age}_i + \beta_{\mathrm{W}}\mathrm{Wealthindex}_i + \beta_{\mathrm{Ls}}\mathrm{Latent}_{i\mathrm{s}} \nonumber\\
\alpha_{0,\mathrm{h}\mathrm{s}} &\sim \mathrm{Normal}(0, 2),~~ h=1, \cdots, \mathrm{H} \nonumber\\
\mathrm{Age}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
\mathrm{Wealthindex}_i &\sim \mathrm{Normal}(0, 1) \nonumber\\
\mathrm{Latent}_{[i]\mathrm{s}} &\sim \mathrm{Normal}(0, 1) \nonumber\\
\beta_0 &= 5 \nonumber\\
\beta_{\mathrm{A}} &= 2 \nonumber\\
\beta_{\mathrm{W}} &= -3 \nonumber\\
\beta_{\mathrm{L\text{\tiny{water}}}} &= 1 \nonumber\\
\beta_{\mathrm{L\text{\tiny{garbage}}}} &= -5 \nonumber\\
i &= 1,\cdots, 100
\end{align}
%
Where:
\begin{itemize}
\item $\beta_{0,s}$ \textbf{is the service specific effect???}, i.e., baseline log odds of a household having improved service
\item $\alpha_{0,\mathrm{h}\mathrm{s}}$ household specific random intercepts, i.e., captures the offset from the overall baseline (mean)
\end{itemize}

@@

\begin{appendix}

\section{Appendix A: Additional simulations}

\subsection{Mediated variables}

We consider simple indirect mediation example such that $x$ has direct effect on $y$ which in turn has effect on $z$ but $x$ has no direct effect on $y$, i.e., $x \rightarrow y \rightarrow z$. In particular:
%
\begin{align}
z &= \beta_0 + \beta_{yz} y + \epsilon_z \nonumber\\
y &= \beta_{xy} x + \epsilon_y \nonumber\\
x &\sim \mathrm{Normal(0, 1)} \nonumber\\
\epsilon_y &\sim \mathrm{Normal(0, 1)} \nonumber\\
\epsilon_z &\sim \mathrm{Normal(0, 1)} \nonumber\\
\beta_0 &= 5 \nonumber\\
\beta_{xy} &= 1 \nonumber\\
\beta_{yz} &= 1.5
\end{align}

%
\begin{figure}[H]
\begin{center}
<<pred_mediate_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
print(pred_mediate_plots)
@
\end{center}
\caption{Predictor effect plot for multivariate response model. The top plot shows the predicted probability of improved service at the baseline. The bottom plots compares the predictor effect of age and wealth index on the expected probability of improved service. The grey points are the observed proportions (or bins in bottom plots) of improved service. The red points and curves are the bias uncorrected estimates while the black points and curves are the population-based (bias corrected) estimates. Generally, without taking into account the potential bias, we risk overestimating the predicted probabilities.}
\label{fig:pred_mediate_plots}
\end{figure}
%

%
\begin{figure}[H]
\begin{center}
<<pred_mediate_bin_plots, echo=FALSE, fig=TRUE, message=FALSE>>=
print(pred_mediate_bin_plots)
@
\end{center}
\caption{Predictor effect plot for multivariate response model. The top plot shows the predicted probability of improved service at the baseline. The bottom plots compares the predictor effect of age and wealth index on the expected probability of improved service. The grey points are the observed proportions (or bins in bottom plots) of improved service. The red points and curves are the bias uncorrected estimates while the black points and curves are the population-based (bias corrected) estimates. Generally, without taking into account the potential bias, we risk overestimating the predicted probabilities.}
\label{fig:pred_mediate_bin_plots}
\end{figure}
%

\end{appendix}
